---
title: "Joining and reshaping data in R"
output: html_notebook
---


Libraries and SELECTION

I usually start off by putting the libraries I'll be using at the top, along with any manually selected items to be filtered (e.g. Math or ELA, Grade selection, etc. . .)
```{r library}
# NOTE: The pound sign -- that's a hashtag to you young'uns -- comments out the rest of the line

library(tidyverse) # this is a selected group of additional tools I find useful 
# tidyverse contains tidyr to reshape with one command.  other library options: reshape, reshape2
library(DBI) # this will help you connect to your SQL server.  Comment it out for now.

##################################################################
#VALUE ENTRY
#
#gr <- '03'   example                                                    
##################################################################
```

```{r DB, eval=FALSE, include=FALSE}
#DATABASES - Connect to SQL SERVER / ODBC 
#replace "AAB-SQL" with your server and "cyee" with your user ID.  It will prompt for pwd

con <- DBI::dbConnect(odbc::odbc(), "AAB-SQL", UID="cyee", PWD    = rstudioapi::askForPassword("Database password"))

```

SAMPLE SQL block.  Note that I've deactivated it in the top right hand corner.  This adds additional language to the space between {}. 
```{sql SQL CUTSCORE, eval=FALSE, connection=con, include=FALSE, output.var="cutscore"}

--to import data from your SQL server to use in R, just copy your SQL code here.  Be sure to use SQL language

SELECT [grade]
      ,[cut_pl2]
      ,[cut_pl3]
      ,[cut_pl4]
      ,[max]
  FROM [Assessment].[dbo].[cutscores]

```

```{r}
#We will be importing data from sample.data.xls using the RStudio Environment interface.
#We will also copy and paste the code into our R block.
#We will also name this block "XLS import" using RStudio "Modify Chunk Options"

```

```{r reshape}
#try the command head(sample_data) to preview your dataset
#Find the key and value columns, then create wide table using spread()
#####################
# Making a wide dataset using spread()
#####################
sample_data_wide <- spread(sample_data, TestSubj, SS_TOT)


#####################
# Making a long dataset using gather()
# for this example we are returning the wide version back to the original
#####################
sample_data_gather <- gather(sample_data_wide, TestSubj, SS_TOT, E:M, factor_key=TRUE)

# The arguments to gather():
# - data: Data object
# - key: Name of new key column (made from names of data columns)
# - value: Name of new value column
# - ...: Names of source columns that contain values
# - factor_key: Treat the new key column as a factor (instead of character vector)

```

```{r join}
#Let's import complex area names from a csv file using the RStudio Environment
#USE: From Text(readr) option and import ca.map.data.csv
#copy the code below:


#Now, let's join it with our initial sample_data object
#If the names of the two fields to join are the same, it's much easier
#Note, This will merge only common cases for both datasets add all=TRUE to override

sample_data_joined_ca <- merge(sample_data, ca_map_data, by = c("Scode"))

#If the names of the two fields are not the same, you can either rename one of the fields ahead of merging (df stands for data frame)

    ###  rename(df, "newcolumnname"="oldcolumnname")

# OR  specify the match like so

    ### new.df <- merge(df1, df2, by.x=c("df1field"), by.y c("df2field"))

#  As you might have guessed, it's just easier to rename the column ahead of the merge
```

Cleaning up when you're done.  If you play with R enough, you'll end up making some interesting data objects that you're not happy with.  To remove an object, the command is rm()  That's all.  just rm and put the name of the object between the parenthesis.  If you want to remove all of your objects, type rm(list=ls())

Happy exploration, I hope this short tutorial will give you new tools to wrangle data with!
